---
title: "Time Series - Final Project"
author: "by ... (your names)"
date: ' '
output:
  html_document:
    df_print: paged
---

**Remove these hints** :-)

0.  It may be necessary to install the knitr package.
1.  Choose from the menu *Help-\>Markdown Quick Reference* to get information on creating an html/pdf report in R-Studio.\
2.  The report should include *R code* and comments on the results.\
3.  Click the *Knit* button to generate your html/pdf report.

------------------------------------------------------------------------

**You have two tasks below:**

------------------------------------------------------------------------

# TASK 1:

Consider the monthly *numbers of visitors* (see file dane_AJWP.csv at UPeL)).\
1. Make *numbers of visitors* forecasts for each month of the year 2024.

------------------------------------------------------------------------

## SOLUTION of TASK 1

First let's define useful functions.

```{r}
periodogram <- function(TS)
{
  n = length(TS)
  Per= Mod(fft(TS-mean(TS)))^2/n  
  Freq = (1:n -1)/n
  plot(Freq[1:n], Per[1:n], type='h', lwd=2, ylab="Periodogram",
       xlab="Frequency")    
 
}
```

Function periodogram plots the periodogram of time series TS.

```{r}
periods <- function(TS)
{ 
  n = length(TS)
  Per= Mod(fft(TS-mean(TS)))^2/n  #
  Freq = (1:n -1)/n
  nn=n/2
  y = cbind(1:nn, Freq[1:nn], 1/Freq[1:nn],Per[1:nn]); y[order(y[,4],decreasing = TRUE),]
}
```

Function periods shows possible periods.

```{r}
ACF_and_PACF <- function(TS, LagMax)
{
  par(mfrow = c(1, 2))
  acf(na.omit(TS), lag.max = LagMax)
  pacf(na.omit(TS), lag.max = LagMax)
  par(mfrow = c(1, 1))
}
```

Function ACF_and_PAC plots ACF and PACF.

PROPER TASK:

We import the data and plot it.

```{r}
data <- read.csv(file = "dane_AJWP.csv")
TS <- ts(data$numbers.of.V., start = c(2012, 1), frequency = 12)
par(mfrow=c(1,1))
plot(TS)
```

Now we check if the data are seasonal by using periodogram.

```{r}
periodogram(TS)
periods(TS)
```

As we can see, the biggest value in periodogram is for period = 4 months and the second biggest value is for period = 12 months.

We decide to take period = 12 months as it is very typical for this type of data (number of visitors).

We remove seasonality by differentiating and plot the obtained time series.

```{r}
D12_TS <- diff(TS, lag = 12)
plot(D12_TS)
```

We remove the trend.

```{r}
final_TS <- diff(D12_TS, lag = 1)
plot(final_TS)
```

Now we plot the ACF and PACF of our final data.

```{r}
ACF_and_PACF(final_TS, 80)
```

We will try to fit the data to some SARIMA $(p,d,q)×(P,D,Q)_s$ model (with s = 12).

Looking at the ACF and PACF plots we see that:

-   We can try P=1, Q=0 or P=0. Q=1.

-   PACF tails off and ACF cuts off at lag 1, 3 or maybe 4. So we consider p=0 and:

    -   q=1,

    -   q=3,

    -   q=4.

P = 1, Q = 0

```{r}
library(astsa)
for(i in c(1,3,4))
{
  mA<-sarima(TS,0,1,i,1,1,0,12)
  print(mA$ttable)
}
```

P = 0, Q = 1

```{r}
for(i in c(1,3,4))
{
  mA<-sarima(TS,0,1,i,0,1,1,12)
  print(mA$ttable)
}
```

Making predicions.
```{r}
pred<-sarima.for(TS,n.ahead=12,0,1,3,0,1,1,12)
```

------------------------------------------------------------------------

# TASK 2:

Consider the [Wojciech's time series from Lab 1 task 4.2](https://upel.agh.edu.pl/mod/forum/discuss.php?d=8438) (Weekly closing prices from WSE of the PKN Orlen) 1. Make all necessary transformations to get a stationary time series (noise).\
2. If possible, choose the ARMA (p, q) model for stationary noise.

------------------------------------------------------------------------

## SOLUTION of TASK 2

Installs and loads necessary R libraries, including "tidyverse," "tseries," and "forecast."

```{r}
# install.packages(c("tidyverse", "tseries", "conflicted", "forecast"))
library(conflicted)
library(tidyverse)
library(tseries)
library(forecast)
```

Resolves any potential conflicts in function names between loaded libraries using the conflicted package.

```{r}
conflict_prefer("filter", "stats")
conflict_prefer("lag", "stats")
```

Reads a CSV file named "PKN.WA.csv" containing financial data into a data frame called data.

```{r}
data <- read.csv("PKN.WA.csv")
```

Creates a time series object (TS) from the closing prices in the data, specifying a frequency of 52 (assuming weekly data) and a starting date of January 2009.

```{r}
TS <- ts(data$Close, frequency = 52, start = c(2009, 1))
```

Plots the original time series to visualize the closing prices over time.

```{r}
plot(TS, main = "Original Time Series", ylab = "Closing Prices", xlab = "Time")
```

Applies differencing to the time series (TS_diff) to remove the trend.

```{r}
TS_diff <- diff(TS)
plot(TS_diff, main = "Differenced Time Series", ylab = "Differenced Closing Prices", xlab = "Time")
```

Applies seasonal differencing to the differenced time series (TS_seasonal_diff) to remove seasonality.

```{r}
TS_seasonal_diff <- diff(TS_diff, lag = 52)
plot(TS_seasonal_diff, main = "Seasonally Differenced Time Series", ylab = "Seasonally Differenced Closing Prices", xlab = "Time")
```

Performs an Augmented Dickey-Fuller test on the seasonally differenced time series to check for stationarity.

```{r}
adf.test(TS_seasonal_diff)
```

Plots the autocorrelation and partial autocorrelation functions of the seasonally differenced time series.

```{r}
par(mfrow=c(1,2))
acf(TS_seasonal_diff)
pacf(TS_seasonal_diff)
par(mfrow=c(1,1))
```

Applies a logarithmic transformation to the original time series (TS_log) to stabilize variance.

```{r}
TS_log <- log(TS)
```

Applies differencing to the log-transformed time series (TS_log_diff) and performs an Augmented Dickey-Fuller test to check for stationarity.

```{r}
TS_log_diff <- diff(TS_log)
adf.test(TS_log_diff)
```

Plots the final transformed time series after applying logarithmic transformation and differencing.

```{r}
plot(TS_log_diff, main = "Final Transformed Time Series", ylab = "Log Differenced Closing Prices", xlab = "Time")
```

Plots the autocorrelation and partial autocorrelation functions of the log-differenced time series.

```{r}
par(mfrow=c(1,2))
acf(TS_log_diff)
pacf(TS_log_diff)
par(mfrow=c(1,1))
```

Defines an ARIMA model with specified order (p, 0, q) and fits it to the log-differenced time series.

```{r}
p <- 3
q <- 1
arima_model <- arima(TS_log_diff, order = c(p, 0, q))
```

Extracts residuals from the ARIMA model.

```{r}
residuals <- residuals(arima_model)
```

Generates a diagnostic plot for the ARIMA model.

```{r}
tsdiag(arima_model)
```

Prints a summary of the ARIMA model.

```{r}
summary(arima_model)
```

Forecasts future values using the fitted ARIMA model for the next 10 time points.

```{r}
forecast_values <- forecast(arima_model, h = 52)
```

Plots the forecasted values against time.

```{r}
plot(forecast_values, main = "ARIMA Forecast", ylab = "Log Differenced Closing Prices", xlab = "Time")
```

Plots the residuals of the ARIMA model against time.

```{r}
plot(residuals, main = "Residuals", ylab = "Residuals", xlab = "Time")
```
------------------------------------------------------------------------

GOOD LUCK !!!
